---
title: "Your Project Title"
subtitle: "Put subtitle here if you want or remove if not needed"
author: 
  - One Author's Name (REPLACE)
  - Another Author's Name (REPLACE)
abstract: "Executive Summary. <!--For the final version: Write an executive summary here. Write it when you are mostly finished with the main report. It should summarize the question, with what data and how you answer it, and what the result is.-->"
format:   
  html: 
    toc: true
    toc-depth: 3
    standalone: true
    embed-resources: true
    code-fold: true
    number-sections: true
---

## Introduction

The COVID-19 pandemic created significant challenges for global health systems, and it's important to highlight that the impact on mortality was unevenly distributed. This report examines the factors that contribute to the differences in COVID-19 mortality rates among various countries since the beginning of the pandemic.

Research Question: From March 2020 to September 2022, how did COVID-19 mortality levels vary globally, and what relative roles did economic capacity (GDP) and societal demographics (Age, Healthcare) play in driving these differences?

To answer this, we will integrate daily mortality records from the WHO COVID-19 database alongside country-level socioeconomic indicators sourced from the World Bank (WDI). In our evaluation of these specific indicators, we seek to determine whether economic resources served as a protective shield or if demographic vulnerabilities, like an aging population, had a more significant, confounding impact.

## Data 

Our analysis integrates two datasets provided in the project repository:

COVID-19 Data: Daily new cases and deaths from the WHO COVID-19 database, covering the period from March 1, 2020, to September 30, 2022.

Socioeconomic Indicators: Country-level socioeconomic data from the World Bank's WDI, including GDP per capita, population size, elderly population percentage, and physician density.

In the data preparation phase, we standardized variable names across the datasets and aggregated the daily WHO records to derive a cumulative death count for each country. To account for vast differences in population size, we normalized these counts into a "deaths per 100,000 people" metric. The final dataset was created by merging these mortality records with the socioeconomic factors and removing entries with missing values, yielding a robust sample of 168 countries for the statistical analysis.

```{r}
#| label: data_import
#| message: false
#| warning: false

library(tidyverse)
library(knitr)
library(broom)

who_raw <- read_csv("WHO-COVID-19.csv")
wdi_raw <- read_csv("WDI.csv")

# Align variable names with group convention
wdi <- wdi_raw |>
  rename(
    pop_total = `Population, total`,
    urban_pop = `Urban population (% of total population)`,
    pop_older65 = `Population ages 65 and above (% of total population)`,
    physicians_per_1000 = `Physicians (per 1,000 people)`,
    gdp_per_capita = `GDP per capita, PPP (current international $)`,
    life_expectancy = `Life expectancy at birth, total (years)`
  ) |>
  select(country, pop_total, urban_pop, pop_older65, physicians_per_1000, gdp_per_capita)

# Aggregate daily deaths into a cumulative total per country
# Filter limited to Sept 2022 to match dataset range
mortality_df <- who_raw |>
  mutate(Date_reported = as.Date(Date_reported)) |>
  filter(Date_reported >= "2020-03-01" & Date_reported <= "2022-09-30") |>
  group_by(Country) |>
  summarize(
    total_deaths = sum(New_deaths, na.rm = TRUE),
    .groups = "drop"
  )

# Merge datasets and calculate per-capita rate
final_df <- mortality_df |>
  inner_join(wdi, by = c("Country" = "country")) |>
  mutate(deaths_per_100k = (total_deaths / pop_total) * 100000) |>
  filter(!is.na(deaths_per_100k) & !is.na(gdp_per_capita) & !is.na(pop_older65))

head(final_df) |>
  kable(caption = "Preview of Aggregated Mortality Data (2020-2022)")
```

## Analysis

We first examine the relationship between national wealth and pandemic outcomes through visualization. The scatter plot below displays a surprising positive trend (indicated by the red dashed line), suggesting that wealthier nations experienced higher mortality rates. However, this visual correlation is confounded by demographic factors; as the bubble sizes indicate, wealthier countries also possess significantly older populations who were more vulnerable to the virus.

```{r}
#| label: visualization
#| warning: false
#| fig.cap: "GDP per Capita vs. Cumulative COVID-19 Mortality (2020-2022)"

ggplot(final_df, aes(x = gdp_per_capita, y = deaths_per_100k)) +
  geom_point(aes(size = pop_older65, color = physicians_per_1000), alpha = 0.7) +
  geom_smooth(method = "lm", color = "darkred", linetype = "dashed") +
  scale_x_log10(labels = scales::dollar) +
  scale_color_viridis_c(option = "magma", name = "Physicians/1k") +
  labs(
    title = "GDP per Capita vs. COVID-19 Mortality Rate",
    subtitle = "Analysis from March 2020 to Sept 2022",
    x = "GDP per Capita (Log Scale)",
    y = "Total Deaths per 100k People",
    size = "% Pop over 65"
  ) +
  theme_minimal()
```

To resolve this paradox, we fitted a multiple linear regression model to quantify the distinct impacts of GDP, physician density, and age structure. The regression results confirm that demographics are the dominant driver, with the population over age 65 showing a strong, statistically significant positive association with mortality (p < 0.001). Crucially, when controlling for this age factor, the model reveals that higher GDP is actually associated with a significant decrease in mortality (estimate: -0.001, p = 0.004), overturning the initial visual impression. Furthermore, the positive coefficient for physician density suggests a surveillance bias, indicating that countries with greater medical capacity were better at detecting and reporting COVID-19 deaths.
  
```{r}
#| label: analysis

# Multiple linear regression to predict mortality rate
model_mortality <- lm(deaths_per_100k ~ gdp_per_capita + physicians_per_1000 + pop_older65, 
                      data = final_df)

tidy(model_mortality) |>
  kable(
    caption = "Linear Regression Results: Predictors of COVID-19 Mortality",
    digits = 3
  )
```

## Conclusion

This analysis investigates the global disparities in COVID-19 mortality from March 2020 to September 2022, emphasizing the complex relationship between economic capacity and demographic structure. In our statistical modeling, we found that societal demographics, particularly the percentage of the population over age 65, emerged as the primary predictor of mortality variability. It appears that the biological vulnerability of older populations significantly surpasses most economic advantages.

I've noticed that simple visualizations indicate a contradictory trend where wealthier nations seem to have higher death rates, often referred to as the "Reporting Paradox." However, when we control for age, it appears that economic capacity, measured by GDP, actually contributes to a statistically significant, although modest, reduction in mortality rates.

Additionally, the analysis revealed a surveillance effect, showing that a higher density of physicians correlated with increased reported mortality. This suggests that "societal factors" include not only health capacity but also the ability to detect cases, as countries with more doctors were likely better positioned to accurately diagnose and document COVID-19 deaths in comparison to those with fewer resources.

One of the main limitations we face is the consistency of the data. Definitions of "COVID-19 Death" have varied across different countries, and I believe that under-reporting in developing nations continues to be a major source of bias. In future research, it would be beneficial to examine "Excess Mortality" in order to address these reporting gaps.
